\section*{Deviations from JavaScript}

We intend the 
Source language to be a conservative extension
of JavaScript: Every correct
Source program should behave \emph{exactly} the same
using a Source implementation, as it does using a JavaScript
implementation. We assume, of course, that suitable libraries are
used by the JavaScript implementation, to account for the predefined names
of each Source language.

This section lists some exceptions where we think a Source implementation
should be allowed to deviate from the JavaScript specification, for the
sake of internal consistency and esthetics.

\begin{description}
\item[\href{https://source-academy.github.io/sicp/chapters/4.1.1.html\#footnote-4}{Evaluation result of programs:}]
  % from SICP JS 4.1.2, last exercise
JavaScript statically
        distinguishes between \emph{value-producing} and
	\emph{non-value-producing statements}. All declarations are
	non-value-producing, and all 
	expression statements, conditional statements and assignments are
	value-producing.
	A block is value-producing if its body statement
	is value-producing, and then its value is the value of its body
	statement. A sequence is value-producing if any of
	its component statements is value-producing, and then its value is
	the value of its \emph{last} value-producing component statement.
	The value of an expression statement is the value of the expression.
	The value of a conditional statement is the value of the branch that
	gets executed, or the value
	\lstinline{undefined} if that branch is
	not value-producing.
	The value of an assignment is the value of the expression
	to the right of its \lstinline{=} sign.
	Finally, if the whole
	program	is not value-producing, its value is the value
        \lstinline{undefined}.        

  Example 1:
  \begin{lstlisting}
1;
{
  // empty block
}
  \end{lstlisting}
  The result of evaluating this program in JavaScript is 1.

  Example 2:
  \begin{lstlisting}
1;
{
  if (true) {} else {}
}
  \end{lstlisting}
  The result of evaluating this program in JavaScript is \texttt{undefined}.

  Implementations of Source are currently allowed to opt for a simpler scheme. 

\item[\href{https://source-academy.github.io/sicp/chapters/1.3.2.html\#footnote-2}{Hoisting of function declarations:}] In JavaScript, function declarations
  are ``hoisted''
  (\href{https://source-academy.github.io/sicp/chapters/4.3.1.html#footnote-4}{automagically} moved)
  to the beginning of the block in which
  they appear. This means that applications of functions that are declared
  with function declaration statements never fail because the name is not
  yet assigned to their function value. The specification of Source does 
  not include this hoisting; in Source, function declaration can be seen as
  syntactic sugar for constant declaration and lambda expression.
  As a consequence, application of functions declared with function declaration
  may fail in Source if the name that appears as function expression is not yet
  assigned to the function value it is supposed to refer to.
\end{description}
